---
title: "Homework 1"
subtitle: "Advanced Data Programming with R"
author: "Roshini Gopinath"
format:
  html:
    fontsize: 1.0em
    embed-resources: true
    grid:
      body-width: 800px
  revealjs: 
    slide-number: true
    html-math-method: mathjax
    theme: serif
execute: 
  eval: true
  echo: false
  warning: false
  message: false
---

```{r}
# Load required libraries
library(dplyr)
library(knitr)
library(tidyr)
library(ggplot2)
data(diamonds)
library(GGally)
```

# Introduction

The diamonds dataset from the **ggplot2** package. The dataset contains **53,940 observations** of individual diamonds, each described by 10 features .It holds stuff like how heavy they are (carat), how well they’re cut, their color, clarity, and of course, their price.

For this quick analysis, I wanted to see how diamond **`price`** changes with **`carat size`**, and how that relationship shifts depending on the **`cut quality`**. Basically: do better cut diamonds get priced higher even if they’re the same size? Let’s find out.

# *Descriptive Statistics for Numeric Features in the Diamonds Dataset*

```{r}
# Create a descriptive statistics table for selected numeric variables
numeric_summary <- diamonds %>%
  
   # Select only numeric columns of interest
  select(carat, depth, table, price, x, y, z) %>%
  
   # Compute summary statistics: Min, Max, Mean, and Standard Deviation
  summarise_all(list(
    Min = ~min(.),
    Max = ~max(.),
    Mean = ~round(mean(.), 2),
    SD = ~round(sd(.), 2)
  )) %>%
  pivot_longer(cols = everything(), names_to = c("Variable", ".value"), names_sep = "_")

# Display table
kable(numeric_summary, caption = "Descriptive Statistics for Numeric Features in the Diamonds Dataset")

```

-   **Carat** values range from **0.2 to 5.01**, with an average just under **1 carat** . So most diamonds are fairly small.

-   **Price** spans a massive range, from **\$326 to over \$18,000**, with a high standard deviation, showing a lot of variation.

-   The **depth** and **table** percentages are pretty consistent across the dataset, with small standard deviations, but still important in determining cut quality.

-   Dimensions (`x`, `y`, `z`) generally center around **5–6 mm**, though the presence of **zeros** may indicate missing or misrecorded values that should be cleaned if modeling.

# Pairwise Scatterplot Matrix

```{r}

# Sampling a smaller subset of the dataset(because full dataset is huge)
diamonds_small <- diamonds[sample(nrow(diamonds), 500), ]

# Selecting only the numeric variables for the matrix plot
numeric_vars <- diamonds_small[, c("carat", "depth", "table", "price", "x", "y", "z")]

# Generate a pairwise scatterplot matrix of selected numeric features
ggpairs(numeric_vars,
        title = "Pairwise Scatterplot Matrix of Diamond Features")


```

## Observations

This pairwise scatterplot matrix gives a full view of how the diamond’s physical characteristics relate to one another and to price.

-   **Carat** stands out with a strong positive correlation with **price**, as expected ,heavier diamonds cost more.There's a clear link between **carat** and the physical dimensions (x, y, z), since bigger stones take up more space.

-   Variables like **depth** and **table** don’t show strong correlations with price ,their impact seems more subtle.

# **Scatter Plot of Diamond Price vs Carat Size**

```{r}
# Set seed for reproducibility 
set.seed(123)
# Randomly sample 500 diamonds from the full dataset to improve plotting performance
diamonds_small <- diamonds[sample(nrow(diamonds), 500), ]

# Create a scatter plot of Price vs Carat, coloured by Cut
ggplot(diamonds_small, aes(x = carat, y = price, color = cut)) +
  geom_point(alpha = 0.6) +
  theme_minimal() +
  labs(title = "Diamond Price vs Carat Size",
       subtitle = "Coloured by Cut Quality",
       x = "Carat",
       y = "Price ($)")
```

## Observations

-   **Fair** diamonds tend to appear at higher carat sizes, which likely explains their high average price despite being the lowest cut quality.

-   **Good** and **Very Good** cuts are spread across mid-range carat and price values, showing a balance between size and quality.

-   **Premium** cuts show up across a wide range of carat sizes and consistently command solid prices.

-   **Ideal** cuts dominate the lower carat range but still often reach high price points, proving that excellent cut quality increases value even in smaller stones.

    In summary: **Carat influences the base price**, but **cut quality drives the final value**, introducing significant price differences among diamonds of similar size.
